---
title: "Part 2: Data Analysis"
author: "David Murillo"
date: 'December 4, 2022'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# complete data analysis on the Delomys species data {.tabset .tabset-pills}

## Data Exploration

### Numerical Exploration

```{r, include= FALSE}
require(here)
delomys <- read.csv(here("data", "delomys.csv"))

```

Create a code chunk that includes the following:

- Use summary() on the body mass and body length data columns in the Delomys data set to display summary statistics.

```{r}
summary(delomys[, c("body_mass", "body_length")])
```

- Perform a test of normality on the body mass and length columns. You can use shapiro.test()

```{r}
shapiro.test(delomys$body_mass)

shapiro.test(delomys$body_length)
```

### Graphical Exploration

Using code chunks, create the following plots, which you’ll use to answer the report questions:

- A scatterplot of body mass and body length

```{r, fig.align='center', fig.width= 6, fig.height= 4, echo=FALSE}
plot(body_mass ~ body_length, data= delomys,
     main = "Scatterplot of body mass with body length",
     ylab = "Body length (mm)",
     xlab = "Body length (mm)",
     pch = 20,
     cex= 1.5)
```


- A histogram of body mass

```{r, fig.align='center', fig.width= 6, fig.height= 4, echo=FALSE}
hist(delomys$body_mass,
     main = "Histogram of body mass",
     xlab = "Body length (mm)",
     breaks = 20,
     xlim = c(0,120),
     ylim = c(0, 250))
```

- A histogram of body length

```{r, fig.align='center', fig.width= 6, fig.height= 4, echo=FALSE}
hist(delomys$body_length,
     main = "Histogram of body length",
     xlab = "Body length (mm)",
     breaks = 20,
     xlim = c(40,170),
     ylim = c(0, 550))
```


- A conditional boxplot of body mass, conditioned on species (column binomial)

```{r, fig.align='center', fig.width= 6, fig.height= 4, echo=FALSE}

boxplot(body_mass ~ binomial, data= delomys,
        main = "Body mass by species",
        xlab = "Species",
        ylab = "Body mass (mm)")

```

- A conditional boxplot of body mass, conditioned on sex (column binomial)

```{r, fig.align='center', fig.width= 6, fig.height= 4, echo=FALSE}

boxplot(body_mass ~ sex, data= delomys,
        main = "Body mass by sex",
        xlab = "Sex",
        ylab = "Body mass (mm)")

```
- A conditional boxplot of body mass, conditioned on both species and sex

```{r, fig.align='center', fig.width= 9, fig.height= 6, echo=FALSE}

boxplot(body_mass ~ binomial + sex, data= delomys,
        main = "Body mass by species and sex",
        xlab = "",
        ylab = "Body mass (mm)",
        names = c("Delomys dorsalis\n female", "Delomys sublineatus\n female","Delomys dorsalis\n male","Delomys sublineatus\n male"))

```

### Questions

**Q1 (2 pts.): Qualitatively describe the relationship between body mass and length.**

**Does the relationship seem linear, curved, nonexistent?**

*Answer: *

**Q2 (2 pts.): Qualitatively describe the shapes of the histograms.**

**Do the data appear normally-distributed? Explain why or why not.**

*Answer: *

**Q3 (3 pts.): Using both the histograms and normality tests, do you think the (unconditioned) body masses and body length are normally-distributed?**

**Contrast your visual assessment of normality to the results of the numerical normality tests.**

*Answer: *

**Q4 (2 pts.): Examine the conditional boxplots. Describe any graphical evidence you see for body mass differences based on species and/or sex.**

*Answer: *



## Model Building

Using a code chunk, fit 5 models using lm():

- Model 1: simple linear regression body_length ~ body_mass
- Model 2: 1-way ANOVA body_mass ~ sex
- Model 3: 1-way ANOVA body_mass ~ binomial
- Model 4: 2-way additive ANOVA body_mass ~ sex + binomial
- Model 5: 2-way factorial ANOVA body_mass ~ sex * binomial

```{r}
fit1 <- lm(body_length ~ body_mass, data= delomys)
fit2 <- lm(body_mass ~ sex, data= delomys)
fit3 <- lm(body_mass ~ binomial, data= delomys)
fit4 <- lm(body_mass ~ sex + binomial, data= delomys)
fit5 <- lm(body_mass ~ sex * binomial, data= delomys)
```


## Model Diagnostics

Let’s check whether our models fulfill the assumption of normality of the residuals.

First, use a graphical approach: plot histograms of the model residuals.

```{r}
par(mfrow = c(2,3))

hist(residuals(fit1), 
     breaks = 20, 
     main = "Model fit1",
     xlab = "Residuals")

hist(residuals(fit2), 
     breaks = 20, 
     main = "Model fit2",
     xlab = "Residuals")

hist(residuals(fit3), 
     breaks = 20, 
     main = "Model fit3",
     xlab = "Residuals")

hist(residuals(fit4), 
     breaks = 20, 
     main = "Model fit4",
     xlab = "Residuals")

hist(residuals(fit5), 
     breaks = 20, 
     main = "Model fit5",
     xlab = "Residuals")
```


Next, use shapiro.test() on each model to test the null hypothesis that the residuals are drawn from a normally-distributed population.

```{r}
shapiro.test(residuals(fit1))
shapiro.test(residuals(fit2))
shapiro.test(residuals(fit3))
shapiro.test(residuals(fit4))
shapiro.test(residuals(fit5))
```


### Questions

**Q5 (2 pts.): What do you conclude about residual normality based on the numerical and graphical diagnostics?**

*Answer: *

**Q6 (1 pt.): Are violations of the normality assumption equally severe for all the models?**

*Answer: *

## Model Interpretation

You can use the following code within a code chunk to print out a nicely formatted model coefficient table:

knitr::kable(coef(summary(my_model_fit)))

where my_model_fit is the name of your fitted model object.

You can use similar syntax to print a nicely formatted ANOVA table: knitr::kable(anova(my_model_fit))

Check out the digits argument to control how many decimal digits are printed.

### Body Length 

Print the model coefficient table using summary() and answer the following:

```{r}
knitr::kable(coef(summary(fit1)), digits = 3)
```


**Q7 (2 pts.): What is the magnitude of the mass/length relationship?**

*Answer: *

**Q8 (2 pts.): What is the expected body length of an an animal that weighs 100g?**

*Answer: *

**Q9 (2 pts.): What is the expected body length of an animal that weighs 0g?**

*Answer: *

### Body Mass: Model Coefficients

Print the model coefficient tables for each of the body mass model fits.

```{r}
knitr::kable(coef(summary(fit2)), digits = 3)
knitr::kable(coef(summary(fit3)), digits = 3)
knitr::kable(coef(summary(fit4)), digits = 3)
knitr::kable(coef(summary(fit5)), digits = 3)
```


**Q10 (1 pt.): What is the base level for sex?**

*Answer: *

**Q11 (1 pt.): What is the base level for binomial?**

*Answer; *

**Q12 (1 pt.): Which sex is heavier?**

*Answer: *

**Q13 (1 pt.): Which species is heavier?**

*Answer: *

### Body Mass: ANOVA

Print the ANOVA tables for each of the body mass models.

```{r}
knitr::kable(anova(fit2), digits = 3)
knitr::kable(anova(fit3), digits = 3)
knitr::kable(anova(fit4), digits = 3)
knitr::kable(anova(fit5), digits = 3)
```


**Q14 (1 pt.): Are sex and species significant predictors for body mass?**

*Answer: *

**Q15 (1 pt.): Is there a significant interaction?**

*Answer: *

**Q16 (2 pts.): How dos the significance (as measured by p-value) of either of the main effects (sex and species) differ between the single-predictor models, the additive model, and the interactive model?**

*Answer: *

## Model Comparison: Body Mass

You built four different models of body mass. How do you choose the best one?

One option is to choose the model with the lowest AIC. You can calculate AIC using the appropriately named AIC() function.

Create a code chunk that calculates the AIC values for each of the body mass models.

```{r}
knitr::kable( AIC(fit1, fit2, fit3, fit4, fit5))
```


**Q17 (1 pt.): Which two models have the lowest AIC?**

*Answer: *

**Q18 (4 pts.): Which of the two models with lowest AIC scores would you select?**

**Explain your decision based on model fit and the complexity/understanding tradeoff.**

*Answer: *